Confirmation of Wide-Field Signatures in Redshifted 21~cm Power Spectra using Murchison Widefield Array Observations

### 2015-04-29 ms.tex
Add a couple of justifications, changes in wording
From 02433ee:
============

In ms.tex,

(*) Add justification for using northward baselines for pitchfork demo
(*) Add justification for stopping short of a full quantitative
comparison between data and model
(*) Changes in wording throughout

Here we confirm a recently predicted and previously unknown foreground signature in the 3D power spectra of high redshift 21~cm measurements, wherein the interferometer is sensitive to large-scale structure on all baselines. This is due to the inherently chromatic nature of a wide-field instrument response and is characterized by enhanced power from foreground emission in Fourier modes adjacent to those considered to be most sensitive to the cosmological H~{\sc i} signal. Thus it is a critical input to design and analysis choices of future instruments such as the Hydrogen Epoch of Reionization Array and the Square Kilometre Array. The simulation which predicted this feature was recently validated against Murchison Widefield Array data but this key element was at or below the noise level. In this paper, we improve the Murchison Widefield Array data sensitivity through coherent averaging of 12 independent snapshots aligned in local sidereal time across different observing nights, and provide the first confirmation of the prediction with a signal-noise ratio $>10$.

@@

The epoch of reionization (EoR) commenced following the formation of the first stars and galaxies. It is characterized by a period of non-linear growth of matter density perturbations and astrophysical evolution in the Universe's history. Detection of redshifted 21~cm radiation of H~{\sc i} from this epoch is one of the most promising probes of the evolution of large scale structure characteristic of this epoch \citep{sun72,sco90,mad97,toz00,ili02}.

@@

The primary challenge to detection of cosmological H~{\sc i} from the EoR arises from continuum emission from Galactic and extragalactic foreground objects, relative to which the desired signal is $\sim 10^4$ times weaker. However, the inherent differences in spatial isotropy and spectral smoothness can be exploited to extract the cosmological signal from foreground contamination \citep[see, e.g.,][]{dim02,dim04,zal04,fur04,mor04,san05,fur06,mcq06,mor06,wan06,gle08}. Thus, a detailed characterization of foreground emission has become essential \citep{ali08,bow09,liu09,ber09,ber10,dat10,liu11,gho12,mor12,par12b,tro12,pob13,dil13,dil14,liu14a,liu14b,thy13,thy15}.

@@

A careful design of antenna aperture can significantly mitigate this contamination. Optimal weighting of foreground contaminated Fourier modes may be required to extract the signal with maximum sensitivity. Thus, knowledge of such detailed foreground signatures becomes essential for design and analysis choices of future instruments such as HERA and SKA.

@@

Paper~I describes in detail the effects of wide-field measurements as seen in the delay spectra of interferometer {\it visibilities}. Here, we list the relevant equations and give a brief overview of the wide-field signatures predicted therein.

The delay spectrum for a baseline vector, $\boldsymbol{b}$, is given by \citep[][Paper~I]{par12a,par12b,thy13}: 
\begin{align}
  \tilde{V}_b(\tau) &\equiv \int V_b(f)\,W(f)\,e^{i2\pi f\tau}\,\dif f,
\end{align}
with interferometer visibilities, $V_b(f)$, given by \citep{van34,zer38,tho01}:
\begin{align} 
  V_b(f) &= \iint\limits_\textrm{sky} A(\hat{\boldsymbol{s}},f)\,I(\hat{\boldsymbol{s}},f)\,W_\textrm{i}(f)\,e^{-i2\pi f\frac{\boldsymbol{b}\cdot\hat{\boldsymbol{s}}}{c}}\,\dif\Omega \\
         &= \iint\limits_\textrm{sky} \frac{A(\hat{\boldsymbol{s}},f)\,I(\hat{\boldsymbol{s}},f)}{\sqrt{1-l^2-m^2}}\,W_\textrm{i}(f)\,e^{-i2\pi f\frac{\boldsymbol{b}\cdot\hat{\boldsymbol{s}}}{c}}\,\dif l\,\dif m, 
\end{align}
where, $I(\hat{\boldsymbol{s}},f)$ and $A(\hat{\boldsymbol{s}},f)$ are the sky brightness and antenna's directional power pattern, respectively, as a function of frequency ($f$) and direction on the sky denoted by the unit vector $\hat{\boldsymbol{s}}\equiv (l,m,n)$, $W_\textrm{i}(f)$ denotes instrumental bandpass weights, $W(f)$ is a spectral weighting function that controls the transfer function in the delay transform, $\dif\Omega=(1-l^2-m^2)^{-1/2}\,\dif l\,\dif m$ is the solid angle element to which $\hat{\boldsymbol{s}}$ is the unit normal vector, and $c$ is the speed of light. $\tau=\boldsymbol{b}\cdot\hat{\boldsymbol{s}}/c$ is the geometric delay between antenna pairs measured relative to the zenith and provides a mapping to position on the sky.

The delay power spectrum is defined as \citep[][Paper~I]{par12a}:
\begin{align}\label{eqn:delay-power-spectrum}
  P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel) &\equiv |\tilde{V}_b(\tau)|^2\left(\frac{A_\textrm{e}}{\lambda^2\Delta B}\right)\left(\frac{D^2\Delta D}{\Delta B}\right)\left(\frac{\lambda^2}{2k_\textrm{B}}\right)^2,
\end{align}
with
\begin{align}
  \boldsymbol{k}_\perp &\equiv \frac{2\pi(\frac{\boldsymbol{b}}{\lambda})}{D}, \\
  k_\parallel &\equiv \frac{2\pi\tau\,f_{21}H_0\,E(z)}{c(1+z)^2}, 
\end{align}
where, $A_\textrm{e}$ is the effective area of the antenna, $\Delta B$ is the bandwidth, $\lambda$ is the wavelength of the band center, $k_\textrm{B}$ is the Boltzmann constant, $f_{21}$ is the rest frequency of the 21~cm radiation of H~{\sc i}, $z$ is the redshift, $D\equiv D(z)$ is the transverse comoving distance, $\Delta D$ is the comoving depth along the line of sight, and $h$, $H_0$ and $E(z)\equiv [\Omega_\textrm{M}(1+z)^3+\Omega_\textrm{k}(1+z)^2+\Omega_\Lambda]^{1/2}$ are standard cosmology terms. In this paper, we use $\Omega_\textrm{M}=0.27$, $\Omega_\Lambda=0.73$, $\Omega_\textrm{K}=1-\Omega_\textrm{M}-\Omega_\Lambda$, and $H_0=100\,$km$\,$s$^{-1}\,$Mpc$^{-1}$. $P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel)$ is in units of K$^2$(Mpc/$h$)$^3$.

The defining characteristics of the {\it pitchfork} signature are understood as follows. The steep rise in subtended solid angle near the horizon for a fixed delay bin size significantly enhances the integrated emission near the horizon delay limits in wide-field measurements. This is found to be true for diffuse emission even on wide antenna spacings because their foreshortening towards the horizon makes them sensitive to large angular scales that match the inverse of their foreshortened lengths. In this paper, we confirm this key feature using deeper data from the MWA.

\section{The Murchison Widefield Array Observations}\label{sec:MWA}

The MWA instrument configuration, EoR observing strategy, and analysis procedure applied to individual snapshots used in this study are already described in Paper~I and references therein. In order to reduce thermal fluctuations while maintaining coherence, it is essential to average independent data sets obtained over the same region of sky with identical beamformer settings. Hence, we select a subset of MWA snapshots each of duration 112~seconds obtained over different nights which are aligned to within 72~seconds of each other in {\it local sidereal time} (LST) around a mean LST of 0.04~hours with the MWA tile beam pointed at zenith. The database contains 14 snapshots satisfying these criteria. Two of these snapshots were found to contain amplitude and phase artifacts for a significant duration across different baselines. Hence, they have been excluded from our analysis.

The delay spectra of the rest of the snapshots were verified to be coherent in their amplitudes and phases. These complex valued delay spectra from independent snapshots are averaged together to lower thermal fluctuations without losing coherence from foreground contributions. The results are discussed below.

@@

Figure~\ref{fig:delay-spectra} shows the delay spectra obtained from a single snapshot of MWA data (top), averaging LST aligned delay spectra from 12 individual snapshots from MWA observations on different nights (middle), and from modeling with no thermal noise fluctuations shown for reference (bottom). In all panels, the {\it foreground wedge} bounded by horizon limits (white dotted lines) is prominent. The bright branch of power at $\tau=0$ corresponds to foreground emission from the main lobe of the antenna power pattern pointed at the zenith.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{multi_baseline_fhd_sim_visibilities_amplitudes_comparison_185.0_MHz_30.7_MHz.eps}
\caption{Delay power spectra obtained from a single snapshot ({\it top}), by averaging 12 snapshots of LST aligned MWA data ({\it middle}), and from modeling with no thermal noise added ({\it bottom}). The $x$-axis, denoted by $|\boldsymbol{b}|$ (and $k_\perp$), represents angular (and spatial) scales in the plane of the sky while the $y$-axis, shown in $\tau$ and $k_\parallel$, denotes the spatial scales along the line of sight. White dotted lines are the horizon delay limits. Dynamic range in the delay power spectra of MWA data has increased by a factor $\sim 10$ after averaging (middle) relative to that in a single snapshot (top). Power near the horizon limits caused by wide-field effects are prominent. Faint horizontal features at $\tau=\pm 0.78\,\mu$s are visible due to effective lowering of thermal fluctuations and are the response to periodic coarse band edge flagging of MWA data every 1.28~MHz. \label{fig:delay-spectra}}
\end{figure}

The key defining feature of the {\it pitchfork} signature is an enhancement in foreground power near the horizon limits. In the individual snapshot, similar to the one used in our original study, faint features are visible near the horizon limits. But the high level of thermal fluctuations makes their significance marginal. In contrast, the dynamic range in the averaged data (middle) is a factor $\gtrsim 10$ higher (in delay power spectrum) relative to that in a single snapshot (top), and is consistent with the improvement expected from averaging 12 independent snapshots. With this improvement in sensitivity, the foreground power near the horizon limits (white dotted lines) appears $\gtrsim 10$ times more prominent. Faint horizontal features appear at $\tau = \pm 0.78\,\mu$s also as a result of lowering thermal fluctuations which is not seen in the single snapshot, thus confirming effective lowering of thermal fluctuations. We identify these faint features as the response in delay space of the MWA coarse band edges flagged periodically every 1.28~MHz.

Figure~\ref{fig:3-baseline-comparison-delay-spectra} shows the averaged delay power spectra on three selected baseline vectors oriented northward. Data and noiseless models are shown in black and red respectively. The horizontal dotted black line denotes {\it rms} of thermal fluctuations estimated from data. The vertical dashed line denotes horizon delay limits, and the vertical dot-dashed lines denote delays at which the responses to coarse band edge flagging are expected.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{3_baseline_comparison_CLEAN_fhd_avg_visibilities_amplitudes_185.0_MHz_30.7_MHz.eps}
\caption{Delay power spectra on three antenna spacings oriented northward, obtained by coherent averaging of 12 snapshots aligned in LST. The averaged data and models are shown in black and red respectively. The antenna spacings are 83.5~m ({\it top}), 104.2~m ({\it middle}), and 171~m ({\it bottom}). The horizontal dotted line is the {\it rms} of thermal fluctuations. The vertical dashed lines denote the horizon delay limits. The vertical dot-dashed lines at $\tau=\pm 0.78\,\mu$s correspond to grating responses of periodic flagging of bandpass at intervals of 1.28~MHz. The peaks close to the horizon delay limits are distinctly visible at $\sim 10$-1000~$\sigma$ levels. Differences between model and data are primarily attributed to uncertainties in the foreground model and the MWA tile power pattern. \label{fig:3-baseline-comparison-delay-spectra}}
\end{figure}

The delay power spectra morphologies from data and modeling are remarkably similar in broad shape although they exhibit some differences in the amplitude scales. We attribute these differences to uncertainties in the foreground model, the MWA tile power pattern, thermal fluctuations, and other uncertainties noted in Paper~I.

We focus on the peaks near the horizon limits in the data. Typically, the power near the negative horizon limit is seen with a signal-noise ratio (SNR) $\sim$~10-100, while that around the positive horizon limit is $\sim$~100-1000. Based on the morphological agreement between the data and the models, we conclude the features noted in our current analysis are a robust detection of the {\it pitchfork} signature predicted in our earlier study.

@@

Using deeper MWA data, we have confirmed with high significance the earlier prediction that wide-field EoR measurements suffer significant foreground contamination from near the horizon. This has important implications for instrument design and data analysis of future instruments such as HERA and SKA. For instance, SKA plans to use vertical log-periodic dipole arrays while HERA will use dishes that limit response near the horizon. Details of both telescopes are being finalized.

Precise modeling is thus required to gain a complete understanding of the characteristics of the cosmological signal and the foregrounds. Extending the footprint of surveys matching the frequency and angular resolution of observation such as the MWA Commissioning Survey \citep[MWACS;][]{hur14} and the Galactic and Extragalactic MWA Survey (GLEAM; Wayth et al. submitted) to cover the entire hemisphere, and a detailed model of the instrument will further improve the accuracy of our models. In our earlier study, we proposed a selective flagging of data on different antenna spacings that can potentially mitigate foreground contamination by two orders of magnitude. Following the confirmation presented here, efforts are underway to incorporate this proposed foreground mitigation technique into the MWA data analysis algorithms.

>>>>

Here we confirm a recently predicted and previously unknown foreground signature in the 3D power spectra of high-redshift 21~cm measurements, wherein the interferometer is sensitive to large-scale structure on all baselines. This is due to the inherently chromatic nature of a wide-field instrument response and is characterized by enhanced power from foreground emission in Fourier modes adjacent to those considered to be most sensitive to the cosmological H~{\sc i} signal. Thus it is a critical input to design and analysis choices of future instruments such as the Hydrogen Epoch of Reionization Array and the Square Kilometre Array. The simulation which predicted this feature was recently validated against Murchison Widefield Array data but this key element was at or below the noise level. In this paper, we improve the Murchison Widefield Array data sensitivity through coherent averaging of 12 independent snapshots aligned in local sidereal time across different observing nights, and provide the first confirmation of the prediction with a signal-noise ratio $>10$.

@@

The epoch of reionization (EoR) commenced following the formation of the first stars and galaxies. It is characterized by a period of non-linear growth of matter density perturbations and astrophysical evolution in the Universe's history. Detection of redshifted 21~cm radiation of H~{\sc i} from this epoch is one of the most promising probes of the evolution of large scale structure during this epoch \citep{sun72,sco90,mad97,toz00,ili02}.

@@

The primary challenge to detection of cosmological H~{\sc i} from the EoR arises from continuum emission from Galactic and extragalactic foreground objects, which is $\sim 10^4$ stronger than the desired signal. However, the inherent differences in spatial isotropy and spectral smoothness can be exploited to extract the cosmological signal from foreground contamination \citep[see, e.g.,][]{dim02,dim04,zal04,fur04,mor04,san05,fur06,mcq06,mor06,wan06,gle08}. Thus, a detailed characterization of foreground emission has become essential \citep{ali08,bow09,liu09,ber09,ber10,dat10,liu11,gho12,mor12,par12b,tro12,pob13,dil13,dil14,liu14a,liu14b,thy13,thy15}.

@@

A careful design of antenna aperture can significantly mitigate this contamination. Optimal weighting of foreground contaminated Fourier modes may be required to extract the signal with maximum sensitivity. Thus, knowledge of such detailed foreground signatures is key for design and analysis choices of future instruments such as HERA and SKA.

@@

Paper~I describes in detail the effects of wide-field measurements as seen in the delay spectra of interferometer {\it visibilities}. We list the relevant equations and give a brief overview of the wide-field signatures predicted therein.

The delay spectrum for a baseline vector, $\boldsymbol{b}$, is \citep[][Paper~I]{par12a,par12b,thy13}: 
\begin{align}
  \widetilde{V}_b(\tau) &\equiv \int V_b(f)\,W(f)\,e^{i2\pi f\tau}\,\dif f,
\end{align}
with interferometer visibilities, $V_b(f)$, given by \citep{van34,zer38,tho01}:
\begin{align} 
  V_b(f) &= \iint\limits_\textrm{sky} A(\hat{\boldsymbol{s}},f)\,I(\hat{\boldsymbol{s}},f)\,W_\textrm{i}(f)\,e^{-i2\pi f\frac{\boldsymbol{b}\cdot\hat{\boldsymbol{s}}}{c}}\,\dif\Omega \\
         &= \iint\limits_\textrm{sky} \frac{A(\hat{\boldsymbol{s}},f)\,I(\hat{\boldsymbol{s}},f)}{\sqrt{1-l^2-m^2}}\,W_\textrm{i}(f)\,e^{-i2\pi f\frac{\boldsymbol{b}\cdot\hat{\boldsymbol{s}}}{c}}\,\dif l\,\dif m, 
\end{align}
where, $I(\hat{\boldsymbol{s}},f)$ and $A(\hat{\boldsymbol{s}},f)$ are the sky brightness and antenna's directional power pattern, respectively, as a function of frequency ($f$) and direction on the sky denoted by the unit vector $\hat{\boldsymbol{s}}\equiv (l,m,n)$, $W_\textrm{i}(f)$ denotes instrumental bandpass weights, $W(f)$ is a spectral weighting function that controls the transfer function in the delay transform, $\dif\Omega=(1-l^2-m^2)^{-1/2}\,\dif l\,\dif m$ is the solid angle element to which $\hat{\boldsymbol{s}}$ is the unit normal vector, and $c$ is the speed of light. $\tau=\boldsymbol{b}\cdot\hat{\boldsymbol{s}}/c$ is the geometric delay between antenna pairs measured relative to the zenith and provides a mapping to position on the sky.

The delay power spectrum is defined as \citep[][Paper~I]{par12a}:
\begin{align}\label{eqn:delay-power-spectrum}
  P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel) &\equiv |\widetilde{V}_b(\tau)|^2\left(\frac{A_\textrm{e}}{\lambda^2\Delta B}\right)\left(\frac{D^2\Delta D}{\Delta B}\right)\left(\frac{\lambda^2}{2k_\textrm{B}}\right)^2,
\end{align}
with
\begin{align}
  \boldsymbol{k}_\perp &\equiv \frac{2\pi(\frac{\boldsymbol{b}}{\lambda})}{D}, \\
  k_\parallel &\equiv \frac{2\pi\tau\,f_{21}H_0\,E(z)}{c(1+z)^2}, 
\end{align}
where, $A_\textrm{e}$ is the effective area of the antenna, $\Delta B$ is the bandwidth, $\lambda$ is the wavelength of the band center, $k_\textrm{B}$ is the Boltzmann constant, $f_{21}$ is the rest frequency of the 21~cm radiation of H~{\sc i}, $z$ is the redshift, $D\equiv D(z)$ is the transverse comoving distance, $\Delta D$ is the comoving depth along the line of sight, and $h$, $H_0$ and $E(z)\equiv [\Omega_\textrm{M}(1+z)^3+\Omega_\textrm{k}(1+z)^2+\Omega_\Lambda]^{1/2}$ are standard cosmology terms. In this paper, we use $\Omega_\textrm{M}=0.27$, $\Omega_\Lambda=0.73$, $\Omega_\textrm{K}=1-\Omega_\textrm{M}-\Omega_\Lambda$, and $H_0=100\,$km$\,$s$^{-1}\,$Mpc$^{-1}$. $P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel)$ is in units of K$^2$(Mpc/$h$)$^3$.

The defining characteristics of the {\it pitchfork} signature are understood as follows. The steep rise in subtended solid angle near the horizon for a fixed delay bin size significantly enhances the integrated emission near the horizon delay limits in wide-field measurements. This is found to be true for diffuse emission even on wide antenna spacings because their foreshortening towards the horizon makes them sensitive to large angular scales that match the inverse of their foreshortened lengths. We confirm this key feature using deeper MWA data.

\section{MWA Observations}\label{sec:MWA}

The MWA instrument configuration, EoR observing strategy, and analysis procedure applied to individual snapshots used in this study are already described in Paper~I and references therein. In order to reduce thermal fluctuations while maintaining coherence, it is essential to average independent data sets obtained over the same region of sky with identical beamformer settings. Hence, we select a subset of MWA snapshots each of duration 112~seconds obtained over different nights which are aligned to within 72~seconds of each other in {\it local sidereal time} (LST) around a mean LST of 0.04~hours with the MWA tile beam pointed at zenith. The database contains 14 snapshots satisfying these criteria. Two of these snapshots, which contained amplitude and phase artifacts for a significant duration across different baselines, are excluded from our analysis.

The delay spectra of the rest of the snapshots were verified to be coherent in their amplitudes and phases. Results from this coherent averaging are discussed below.

@@

Figure~\ref{fig:delay-spectra} shows the delay spectra obtained from a single snapshot of MWA data (top), averaging LST aligned delay spectra from 12 individual snapshots from MWA observations on different nights (middle), and from modeling with no thermal noise fluctuations shown for reference (bottom). In all panels, the {\it foreground wedge} bounded by horizon limits (white dotted lines) is prominent. The bright horizontal branch of power at $\tau\simeq 0$ corresponds to foreground emission from the main lobe of the antenna power pattern pointed at the zenith.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{multi_baseline_fhd_sim_visibilities_amplitudes_comparison_185.0_MHz_30.7_MHz.eps}
\caption{Delay power spectra obtained from a single snapshot ({\it top}), by averaging 12 snapshots of LST aligned MWA data ({\it middle}), and from modeling with no thermal noise added ({\it bottom}). The $x$-axis, denoted by $|\boldsymbol{b}|$ (and $k_\perp$), represents angular (and spatial) scales in the plane of the sky while the $y$-axis, shown in $\tau$ and $k_\parallel$, denotes the spatial scales along the line of sight. White dotted lines are the horizon delay limits. Dynamic range in the delay power spectra of MWA data has increased by a factor $\sim 10$ after averaging (middle) relative to that in a single snapshot (top). Power near the horizon limits caused by wide-field effects are prominent. Faint horizontal features at $\tau\simeq\pm 0.78\,\mu$s are visible due to effective lowering of thermal fluctuations and are the response to periodic coarse band edge flagging of MWA data every 1.28~MHz. \label{fig:delay-spectra}}
\end{figure}

In the single snapshot (top), similar to the one used in our earlier study, faint features associated with the {\it pitchfork} signature are visible near the horizon limits. But the high level of thermal fluctuations makes their significance marginal. In contrast, the dynamic range in the averaged data (middle) is a factor $\gtrsim 10$ higher (in delay power spectrum) relative to that in a single snapshot, and is consistent with the improvement expected from averaging 12 independent snapshots. Hence, the foreground power near the horizon limits appears $\gtrsim 10$ times more prominent. Also, faint horizontal features, not seen in the single snapshot, appear at $\tau \simeq \pm 0.78\,\mu$s thus confirming the improvement in sensitivity. We identify these faint features as the response of the MWA coarse band edges flagged periodically every 1.28~MHz.

In order to show that low-level ubiquitous diffuse emission is a significant contributor to the {\it pitchfork} signature, contribution from any strong emission from near the Galactic center needs to be minimized. This is best illustrated with northward antenna spacings which map any residual emission from this region (even though the Galactic center has just set over the western horizon) to $\tau\simeq 0$ and thus reduce the impact on higher delay modes (Paper~I). Figure~\ref{fig:3-baseline-comparison-delay-spectra} shows the averaged delay power spectra on three selected baseline vectors oriented northward. Data and noiseless models are shown in black and red respectively. The horizontal dotted black line denotes {\it rms} of thermal fluctuations estimated from data. The vertical dashed line denotes horizon delay limits, and the vertical dotted-dashed lines denote delays at which the responses to coarse band edge flagging are expected.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{3_baseline_comparison_CLEAN_fhd_avg_visibilities_amplitudes_185.0_MHz_30.7_MHz.eps}
\caption{Delay power spectra on three antenna spacings oriented northward, obtained by coherent averaging of 12 snapshots aligned in LST. The averaged data and models are shown in black and red respectively. The antenna spacings are 83.5~m ({\it top}), 104.2~m ({\it middle}), and 171~m ({\it bottom}). The horizontal dotted line is the {\it rms} of thermal fluctuations. The vertical dashed lines denote the horizon delay limits. The vertical dotted-dashed lines at $\tau = \pm 0.78\,\mu$s correspond to grating responses of periodic flagging of bandpass at intervals of 1.28~MHz. The peaks close to the horizon delay limits are distinctly visible at $\sim 10$-1000~$\sigma$ levels. Differences between model and data are primarily attributed to uncertainties in the foreground model and the MWA tile power pattern. \label{fig:3-baseline-comparison-delay-spectra}}
\end{figure}

We focus on the prominent peaks in data near the horizon limits. Typically, the power near the negative horizon limit is seen with a signal-noise ratio $\sim$~10-100, while that around the positive horizon limit is $\sim$~100-1000.

There is a remarkable agreement in broad morphology between the data and the model. However, some differences in the amplitude scales are noted. We attribute these differences to uncertainties in the foreground model, the MWA tile power pattern, thermal fluctuations, and other uncertainties noted in Paper~I. At present, these uncertainties prevent a more thorough quantitative comparison between the model and the data. Extending the footprint of surveys matching the frequency and angular resolution of observation such as the MWA Commissioning Survey \citep[MWACS;][]{hur14} and the Galactic and Extragalactic MWA Survey (GLEAM; Wayth et al. submitted) to cover the entire hemisphere, and a detailed model of the instrument will improve the accuracy of our model.

We note that reducing uncertainties will only change the relative strength of the {\it pitchfork} signature in our model. However, the effects giving rise to this signature are generic to all wide-field measurements. Thus, the extremely high significance detection of foreground emission near the horizon limits is a robust confirmation of the predicted wide-field measurement phenomenon.

@@

Using deeper MWA data, we have confirmed with high significance the earlier prediction that wide-field EoR measurements suffer significant foreground contamination from near the horizon. This has important implications for instrument design and data analysis of future instruments such as HERA and SKA. For instance, SKA plans to use ``stations'' of $\sim$~35~m diameter that act as aperture arrays consisting of pseudo-randomly placed vertical log-periodic dipole arrays. The ``station beam'' will be obtained by a phased addition of the dipole responses. The typical horizon response of these station beams is $\gtrsim$~30~dB lower than at the zenith. HERA will use dishes that limit response near the horizon to $\lesssim -40$~dB, $\sim$~20~dB lower than that of the MWA. Details of both telescopes are being finalized. Precise modeling is thus required to gain a complete understanding of the characteristics of the cosmological signal and the foregrounds.

In our earlier study, we proposed a selective flagging of data on different baselines that can potentially mitigate foreground contamination by two orders of magnitude. Following the confirmation presented here, efforts are underway to incorporate this proposed foreground mitigation technique into the MWA data analysis.
